Constructing Biological Regulatory Networks from Process Hitting models

### 2014-08-29 main.tex
add keywords; final version 1.1
all source files in the same folder (TCS submission)
add TODOs after the review

\input{macros/macros}
\input{macros/macros-ph}
\input{macros/macros-grn}

% Commandes À FAIRE
%\usepackage{color} % Couleurs du texte
%\newcommand{\warnul}[1]{\textcolor{red}{\underline{#1}}}
%\newcommand{\rewriteng}[1]{\textcolor{blue}{#1}}
%\newcommand{\rewriteil}[1]{\rewriteng{\textbf{[#1]}}}
%\newcommand{\rewrite}[1]{\rewriteil{REWRITE: #1}}
%\newcommand{\towrite}[1]{\rewriteil{TOWRITE: #1}}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}
%\newcommand{\new}[1]{\textcolor{darkgreen}{\textbf{[$^\bigstar$New:} #1\textbf{$^\bigstar$]}}}
%\def\updated#1{{\color{magenta}#1}}
%\newcommand{\updatedmax}[1]{{\color{violet}#1}}
%\newcommand{\new}[1]{#1}

@@

\begin{abstract}
\parindent 0.5cm
The Process Hitting (PH) is a recently introduced framework to model concurrent processes.
Its major originality lies in a specific restriction on the causality of actions, which
makes the formal analysis of very large systems tractable.
PH is suitable to model Biological Regulatory Networks (BRNs) with complete or partial
knowledge of cooperations between regulators by defining the most permissive dynamics
with respect to these constraints.

On the other hand, the qualitative modeling of BRNs has been widely addressed using Ren\'e Thomas'
formalism, leading to numerous theoretical work and practical tools to understand emerging behaviors.
%In addition to be able to use the existing results on this formalism,
%obtaining the Thomas representation of a PH model allows to iteratively refine a model.
Linking this representation to PH would not only allow to use the existing results on this formalism,
but also to iteratively refine the model of a system.

@@

@@

\input{parts/intro}

\section{Frameworks}\label{sec:frameworks}
\input{parts/ph}
\input{parts/grn}

\input{parts/infer-ig}
\input{parts/infer-param}
\input{parts/asp}
\input{parts/example}
\input{parts/discussion}

>>>>

\input{macros}
\input{macros-ph}
\input{macros-grn}

% Commandes À FAIRE
\usepackage{color} % Couleurs du texte
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}
%\newcommand{\rewrite}[1]{\rewriteil{REWRITE: #1}}

@@

\begin{abstract}
\parindent 0.5cm
The Process Hitting (PH) is a recently introduced framework to model concurrent processes.
\todo{The reviewer did not understand the following sentence:}
Its major originality lies in a specific restriction on the causality of actions, which
makes the formal analysis of very large systems tractable.
PH is suitable to model Biological Regulatory Networks (BRNs) with complete or partial
knowledge of cooperations between regulators by defining the most permissive dynamics
with respect to these constraints.

\todo{Rather name Kaufman as the creator of BRNs.}
On the other hand, the qualitative modeling of BRNs has been widely addressed using Ren\'e Thomas'
formalism, leading to numerous theoretical work and practical tools to understand emerging behaviors.
%In addition to be able to use the existing results on this formalism,
%obtaining the Thomas representation of a PH model allows to iteratively refine a model.
Linking this representation to PH would not only allow to use the existing results on this formalism,
but also to iteratively refine the model of a system.

@@

\begin{keyword}
qualitative modeling \sep
model abstraction \sep
interaction graph \sep
René Thomas' parameters \sep
answer set programming
\end{keyword}

@@

\input{parts-intro}

\section{Frameworks}\label{sec:frameworks}
\input{parts-ph}
\input{parts-grn}

\input{parts-infer-ig}
\input{parts-infer-param}
\input{parts-asp}
\input{parts-example}
\input{parts-discussion}

### 2014-08-29 parts-grn.tex
add TODOs after the review

@@

@@

@@

\paragraph{Discussion on maximum level restrictions}
The maximum level $l_a$ of a component $a$ is usually chosen equal to the number of outgoing edges of $a$ in the IG, \ie the number of components that $a$ regulates.
Indeed, each discrete expression level in $\segm{0}{l_a}$ stands for an arbitrary set of concentration levels of the protein coded by $a$
and for which it has a certain influence on several other components that it regulates.
Allowing more expression levels makes several levels useless as they represent the same influence of $a$ than some other levels;
on the contrary, reducing the maximum level requires that several thresholds on outgoing edges of $a$ share the same value, which is biologically unlikely.
However, these considerations are usually taken into account when one has to directly build a model in this framework, but are not mandatory.
In the scope of this paper, we allow any number of levels for components, without considering the number of outgoing edges,
as the number of processes in a PH sort is not constrained in any way.
For instance, in the example of \pref{fig:runningBRN},
it would be more relevant to have $l_a = 1$ and $\GRNedgef{a}{-}{1}{b} \in E$, as $a$ has only one outgoing edge,
but the value $l_a = 2$ makes the example more interesting; this is why $\GRNedgef{a}{-}{2}{b} \in E$ can be arbitrarily chosen.

\paragraph{Discussion on unsigned edges}
Unsigned edges allow to model regulations whose trend is more complex than a simple activation or inhibition.
It has to be noted that using unsigned edges in addition to the regular positive and negative ones does not add expressivity to the framework.
Indeed, signs have no impact on the parametrization (\pref{def:param}) or the dynamics (\pref{def:dynamics}).
They do have an impact later in this paper, when admissible parametrizations are enumerated (\pref{ssec:admissible-K}).
However, signs in an IG are useful to understand the trend of a regulation.
PH actions are very atomistic and the inference of the underlying BRN proposed in \pref{sec:infer-IG}
allows to find the global trends of regulations between components,
which may turn out to be difficult to perform by hand.
As signs have no direct influence on the semantics, unsigned edges can be convenient to model an IG with partial knowledge on some regulations.

>>>>

\todo{Name Kaufman as the creator of BRNs.}

@@

\todo{Discuss about the meaning of signs (especially $+$ and $-$) and their link with parameters.
Especially, these signs introduce redundancy (and possibly contradiction) with the parameters.
\begin{itemize}
  \item We can discuss more about the choice we made here: the signs are only informative
    (it eases the read of the IG), but are used as a basis for the parameters enumeration;
  \item The reviewer suggests to remove signs, or only present them as an abstraction of the monotonicity assumption, given the levels and parameters of a BRN;
  \item It is also possible to add the monotonicity assumption in the definition of a BRN to avoid contradictions.
\end{itemize}}

@@

\todo{The notation $\GRNreg{b}$ is poorly chosen; should be replaced by $E^{-1}(b)$ for instance.}

@@

\paragraph{Discussion on maximum level restrictions}
\todo{The reviewer says that the discussion on the number of maximum levels has
already been fully developed in other articles related to Thomas and that it is not
necessary to develop this just to state that a PH may accept any number of levels.}
The maximum level $l_a$ of a component $a$ is usually chosen equal to the number of outgoing edges of $a$ in the IG, \ie the number of components that $a$ regulates.
Indeed, each discrete expression level in $\segm{0}{l_a}$ stands for an arbitrary set of concentration levels of the protein coded by $a$
and for which it has a certain influence on several other components that it regulates.
Allowing more expression levels makes several levels useless as they represent the same influence of $a$ than some other levels;
on the contrary, reducing the maximum level requires that several thresholds on outgoing edges of $a$ share the same value, which is biologically unlikely.
However, these considerations are usually taken into account when one has to directly build a model in this framework, but are not mandatory.
In the scope of this paper, we allow any number of levels for components, without considering the number of outgoing edges,
as the number of processes in a PH sort is not constrained in any way.
For instance, in the example of \pref{fig:runningBRN},
it would be more relevant to have $l_a = 1$ and $\GRNedgef{a}{-}{1}{b} \in E$, as $a$ has only one outgoing edge,
but the value $l_a = 2$ makes the example more interesting; this is why $\GRNedgef{a}{-}{2}{b} \in E$ can be arbitrarily chosen.

\paragraph{Discussion on unsigned edges}
\todo{Discuss the meaning of signs (cf. supra).
Also, the reviewer proposes to move this discussion before in the text,
and to define unsigned IGs before signed IGs as the latter is just a consequence of the monotonicity assumption.}
Unsigned edges allow to model regulations whose trend is more complex than a simple activation or inhibition.
It has to be noted that using unsigned edges in addition to the regular positive and negative ones does not add expressivity to the framework.
Indeed, signs have no impact on the parametrization (\pref{def:param}) or the dynamics (\pref{def:dynamics}).
They do have an impact later in this paper, when admissible parametrizations are enumerated (\pref{ssec:admissible-K}).
However, signs in an IG are useful to understand the trend of a regulation.
PH actions are very atomistic and the inference of the underlying BRN proposed in \pref{sec:infer-IG}
allows to find the global trends of regulations between components,
which may turn out to be difficult to perform by hand.
As signs have no direct influence on the semantics, unsigned edges can be convenient to model an IG with partial knowledge on some regulations.

### 2014-08-29 parts-infer-ig.tex
add TODOs after the review

@@

@@

@@

\begin{example}
The IG inference of the PH of \pref{fig:runningPH-1} gives the
IG in \pref{fig:BRN-inf1}, containing the following edges:
\begin{align*}
  E_+ &= \{\GRNedgef{b}{+}{1}{a}, \GRNedgef{c}{+}{1}{a}, \GRNedgef{a}{+}{1}{a}, \GRNedgef{b}{+}{1}{b}, \GRNedgef{c}{+}{1}{c}\}\\
  E_- &= \{\GRNedgef{a}{-}{2}{b}\} \qquad\qquad\qquad\qquad\qquad
  E_\uns = \emptyset
\end{align*}
This IG is close to the one in \pref{fig:runningBRN} but not equivalent,
as each component has an additional auto-action.
The auto-actions on $b$ and $c$ are the consequence of a global stability
in some configurations: $c$ never evolves, and neither does $b$ when $a_2$ is not active.
The auto-action on $a$ is mainly caused by its multi-valued nature.

The inference of the refined PH of \pref{fig:runningPH-2} gives the same IG.

>>>>

\todo{In this while section: add informal explanations and maybe more examples.\\
Especially:
\begin{itemize}
  \item Intuitively detail the motivations
  \item Give the intuition of the methods (esp. $\focals$)
  \item Insist on the use/motivation of the definitions
\end{itemize}
The reviewer also proposes to use a biological running example (such as $\lambda$-phage).}

@@

\todo{The reviewer says about the following paragraph: “The inference of an
IG from a dynamics is interesting and important in biology. However the
motivations expressed here are not well argued. In particular I don't find
this “common”. Some bibliographical references are required to consolidate
the arguments or the sentence has to be rephrased.”}

@@

\todo{The two following subsections (down to “Interaction Inference”) have a lot of definitions but few examples and not enough informal explanations, especially on their use.
Especially:
\begin{itemize}
  \item What kinds of concept do represent $\focals$?
  \item What is the importance of a strict context?
  \item Add an outline guiding the reader through these subsections.
\end{itemize}}

@@

\begin{example}
The IG inference of the PH of \pref{fig:runningPH-2} gives the
IG in \pref{fig:BRN-inf1}, containing the following edges:
\begin{align*}
  E_+ &= \{\GRNedgef{b}{+}{1}{a}, \GRNedgef{c}{+}{1}{a}, \GRNedgef{a}{+}{1}{a}, \GRNedgef{b}{+}{1}{b}, \GRNedgef{c}{+}{1}{c}\}\\
  E_- &= \{\GRNedgef{a}{-}{2}{b}\} \qquad\qquad\qquad\qquad\qquad
  E_\uns = \emptyset
\end{align*}
This IG is close to the one in \pref{fig:runningBRN} but not equivalent,
as each component has an additional auto-action.
The auto-actions on $b$ and $c$ are the consequence of a global stability
in some configurations: $c$ never evolves, and neither does $b$ when $a_2$ is not active.
The auto-action on $a$ is mainly caused by its multi-valued nature.

The inference of the PH of \pref{fig:runningPH-1}
(without refinement with cooperative sort) gives the same IG.

### 2014-08-29 parts-intro.tex
add TODOs after the review

While the formal checking of dynamical properties is often limited to small networks because of the
state graph explosion, an other major drawback of this framework is the difficulty to
specify Thomas' parameters, especially for large networks.

@@

The PH is suitable to model BRNs with different levels of abstraction in the specification of
cooperations (associated influences) between components.
This allows to model BRNs with a partial knowledge on precise evolution functions for components
by capturing the largest (the most general) dynamics.

>>>>

While the formal checking of dynamical properties is often limited to small networks because of the
state graph explosion, an other major drawback of this framework is the difficulty to
specify Thomas' parameters, especially for large networks.
\todo{The reviewer proposes to nuance this with several papers:
\begin{itemize}
  \item Corblin, F., Tripodi, S., Fanchon, E., Ropers, D., \& Trilling, L. (2009).
    A declarative constraint-based method for analyzing discrete genetic
    regulatory networks. Bio Systems, 98(2), 91104.
  \item Bollman, D., Coln-Reyes, O., \& Orozco, E. (2007). Fixed points in
    discrete models for regulatory genetic networks. EURASIP journal on
    bioinformatics \& systems biology, 2007, 97356.
  \item Delaplace, F., Klaudel, H., Melliti, T., \& Sen, S. (2012). Analysis
    of modular organisation of interaction networks based on asymptotic
    dynamics. In D. Gilbert \& M. Heiner (Eds.), CMSB 2012
  \item Scalable Steady State Analysis of Boolean Biological Regulatory Net-
    works Ay F, Xu F, Kahveci T (2009) Scalable Steady State Analysis
    of Boolean Biological Regulatory Networks. PLoS ONE 4(12): e7992.
\end{itemize}
This can be in turn be nuanced as some papers only focus on rather specific aspects (like steady-states for instance).}

@@

The PH is suitable to model BRNs with different levels of abstraction in the specification of
cooperations (associated influences) between components.
\todo{Explain the notion of “cooperation” (informally) to be understandable until the definition of cooperative sorts.}
This allows to model BRNs with a partial knowledge on precise evolution functions for components
by capturing the largest (the most general) dynamics.

### 2014-08-29 parts-ph.tex
add TODOs after the review

\begin{example}
\pref{fig:runningPH-1} represents a PH $(\PHs,\PHl,\PHa)$ with
$\PHs = \{a,b,c\}$,
$\PHl_a = \{a_0,a_1,a_2\}$,
$\PHl_b = \{b_0, b_1\}$,
$\PHl_c = \{c_0, c_1\}$, and
\begin{align*}
\PHa & = \{
	\PHfrappe{a_2}{b_1}{b_0},
&&  \PHfrappe{b_0}{a_2}{a_1},
&&	\PHfrappe{c_0}{a_2}{a_1},\\
&&& \PHfrappe{b_0}{a_1}{a_0},
&&	\PHfrappe{c_0}{a_1}{a_0},\\
&&& \PHfrappe{b_1}{a_0}{a_1},
&&	\PHfrappe{c_1}{a_0}{a_1},\\
&&& \PHfrappe{b_1}{a_1}{a_2},
&&	\PHfrappe{c_1}{a_1}{a_2} \}\enspace.
\end{align*}
The action $h=\PHfrappe{b_1}{a_1}{a_2}$ is playable in the state
$s = \state{b_1,a_1,c_0}$; and $s\play h=\state{b_1,a_2,c_0}$.

@@

>>>>

\begin{example}
\todo{\pref{fig:runningPH-1} lacks readability (arrows hitting $a$ are too small).}
\pref{fig:runningPH-1} represents a PH $(\PHs,\PHl,\PHa)$ with
$\PHs = \{a,b,c\}$,
$\PHl_a = \{a_0,a_1,a_2\}$,
$\PHl_b = \{b_0, b_1\}$,
$\PHl_c = \{c_0, c_1\}$, and
\begin{align*}
\PHa & = \{
	\PHfrappe{a_2}{b_1}{b_0},
&&  \PHfrappe{b_0}{a_2}{a_1},
&&	\PHfrappe{c_0}{a_2}{a_1},\\
&&& \PHfrappe{b_0}{a_1}{a_0},
&&	\PHfrappe{c_0}{a_1}{a_0},\\
&&& \PHfrappe{b_1}{a_0}{a_1},
&&	\PHfrappe{c_1}{a_0}{a_1},\\
&&& \PHfrappe{b_1}{a_1}{a_2},
&&	\PHfrappe{c_1}{a_1}{a_2} \}\enspace.
\end{align*}
The action $h=\PHfrappe{b_1}{a_1}{a_2}$ is playable in the state
$s = \state{b_1,a_1,c_0}$; and $s\play h=\state{b_1,a_2,c_0}$.

@@

\todo{The reviewer did not understand the following paragraph about the time shift.
He wonders if it is really important to the understanding of the rest of the paper. Three possibilities here:
\begin{enumerate}
  \item Add more (informal? formal?) explanations,
  \item Remove (or substantially reduce) this paragraph,
  \item Use priorities and prioritized cooperative sorts to remove the temporal shift.
\end{enumerate}
The last possibility would be formally the most accurate but also the most time-consuming one.}
